Cannibalism among the supermassive black holes

### 2018-11-26 paper.tex
Started to add black hole luminosity section.

\documentclass[english, apj]{emulateapj}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{array}
\usepackage{rotating}
\usepackage{units}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{amsbsy}
\usepackage{amstext}
\usepackage{graphicx}
\usepackage{url}
\usepackage{babel}
\usepackage{float}
\usepackage{siunitx}
%\sisetup{output-exponent-marker=\ensuremath{\mathrm{e}}}
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref}
\usepackage[all]{hypcap}

@@

\providecommand{\tabularnewline}{\\}

@@

Here we look at the merger history of 13 exemplary galaxies across the galaxy mass spectrum extracted from a cosmological simulation of hierarchical structure formation. We investigate how, after merging with incoming galaxies, SMBHs diffuse into the cores of the hosts and interact with the resident black hole. We show that gravitational interactions of multiple SMBHs are most probable in high-mass galaxies with total mass $10^{12} \msun < M < 10^{13} \msun$. Galaxies with lower masses have too few mergers with SMBH hosting galaxies. Galaxies with higher masses are more extended, making dynamical friction processes less efficient and hence failing to drive SMBHs into the host galaxy core.

@@

\subsection{Overview of simulation}
Our simulations focus on elliptical galaxies with central SMBHs. The galaxies were given a dark matter background potential using the Stone-Ostriker profile (\citet{2015ApJ...806L..28S}), which is a three-parameter potential-density pair, whose quantities such as density, potential, and binding energy can be written in closed form.  The galaxies were evolved from $4 > z > 0$.  Orbiting black holes were periodically introduced into the "host" galaxy and their dynamical interaction with the background potential and central SMBH were followed, as described further below.

@@

The primary data we extracted from the above work was organized as follows:
\begin{itemize}
    \item Galactic properties - Each galaxy was distinguished by an id number, with properties such as stellar mass, dark matter mass, central black hole id number, and orbiting black hole id numbers, recorded at 37 redshift slices between $4 > z > 0$.
    \item Black Holes properties - Each black hole was also distinguished by an id number, with properties such as seed mass, accreted mass, its host galaxy id number, and time after $z=4$ at which the black hole entered its host galaxy, if it was not the central black hole.
\end{itemize}

@@

We also compared the stellar masses from Lackner12 to  observational and abundance matching data from \citet{2018AstL...44....8K}. Figure \ref{fig:stellar1} is a partial, approximate, replication of their Figure 11 (abundance matching line).  We rescaled our stellar masses at $z=0$ in a lognormal manner about the line in order to match the observed data.  Table \ref{table:gal_char} shows the original and rescaled stellar masses.  In some cases, the original masses were up to 30-40 times larger than the rescaled values, which would have large implications for the ability of the orbiting black holes to reach the center of their host galaxies.

@@

\begin{table}[htbp]
\centering
\caption{Galaxy characteristics at z=0}
\begin{tabular} {|c | c | c| c|}
\hline
$M_{DM}$ & $M_{CBH}$ & $M_{*} (orig.)$ & $M_{*} (rescaled)$ \\
\hline
 \num{9.18e13} 	&	 \num{2.19e9} 	&	\num{1.75e13} 	&	\num{5.14e11}  \\
 \num{7.75e13} 	&	 \num{5.27e9}	&	\num{1.52e13} 	&	\num{4.00e11}  \\
 \num{2.85e13} 	&	 \num{1.46e9}	&	\num{5.58e12} 	&	\num{3.30e11}  \\
 \num{2.48e13} 	&	 \num{2.40e9}	&	\num{4.80e12} 	&	\num{2.25e11}  \\
 \num{2.29e13} 	&	 \num{1.22e9}	&	\num{4.46e12} 	&	\num{2.68e11}  \\
 \num{1.18e13} 	&	 \num{3.62e9}	&	\num{2.32e12} 	&	\num{2.83e11}  \\
 \num{1.03e13} 	&	 \num{9.85e9}	&	\num{1.96e12} 	&	\num{1.62e11}  \\
 \num{1.00e13} 	&	 \num{9.32e9}	&	\num{1.95e12} 	&	\num{2.49e11}  \\
 \num{8.92e12} 	&	 \num{5.60e9}	&	\num{1.74e12} 	&	\num{3.20e11}  \\
 \num{8.56e12} 	&	 \num{6.86e9}	&	\num{1.62e12} 	&	\num{1.83e11}  \\
 \num{6.94e12} 	&	 \num{2.02e9}	&	\num{1.72e12} 	&	\num{1.51e11}  \\
 \num{6.29e12} 	&	 \num{1.22e9}	&	\num{1.23e12} 	&	\num{1.87e11}  \\
 \num{1.84e12} 	&	 \num{2.21e9}	&	\num{5.52e11} 	&	\num{6.42e10}  \\
\hline
\end{tabular}
\label{table:gal_char}
\end{table}

@@

At each iteration in the code, $r_h$ is updated according to the stellar mass and Equations \ref{re} and \ref{sig}.  The core radius, $r_c$, is recalculated only if an orbiting black hole gets within $r_c$.  The work done by diffusion, as described in Section \ref{psd} is calculated, the total potential energy is updated, and $r_c$ is solved for from Equation 8 in Stone-Ostriker.

@@

We can identify Equation \ref{eq:df} as the dynamical friction term. The second term introduces a variance of the friction term, and even allows the SBHs to be accelerated when the velocity of a SBH gets sufficiently small. The third term introduces a change in velocity perpendicular to the direction of motion of the SBH. It is a randomly oriented vector, and hence causes the SBHs to execute a random walk in phase space. The last two terms will establish that the SBHs are ultimately in energy equipartition with the background stars.

@@

\subsubsection{Gravitational wave recoils}
The code \textsc{AR-Chain} includes post-Newtonian terms up to order 2.5. The SMBHs can therefore merge via gravitational wave emission. We include gravitational wave recoils following the prescription outlined in Kulier15, which is based on the fitting formula by \citet{2012PhRvD..85h4015L}. To save computational time, we assume that a merger will be inevitable when the separation between two SBHs becomes smaller than 1.0 Schwarzschild radius. At the moment of the merger, we assume that the spin vectors of the two SBHs are randomly aligned.

@@

\subsection{Simulation setup}
Using the merger tree data, orbiting black holes were injected into its host galaxy at redshift z, at a distance from the galactic center of $R_{e}$ (Eqn. \ref{re}).  Their initial velocity was arbitrarily chosen to be circular, $v_c = \sqrt{M(R_e)/R_e}$, with $v_x$, $v_y$, and $v_z$ randomly chosen.  Note: we only inject the BHs into the simulation that have a $t_{fric}$ smaller than 100 times the Hubble time.

>>>>

\documentclass[english, apj]{emulateapj}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{array}
\usepackage{rotating}
\usepackage{units}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{amsbsy}
\usepackage{amstext}
\usepackage{graphicx}
\usepackage{url}
\usepackage{babel}
\usepackage{float}
\usepackage{savesym}
\savesymbol{tablenum}
\usepackage{siunitx}
\restoresymbol{SIX}{tablenum}
%\sisetup{output-exponent-marker=\ensuremath{\mathrm{e}}}
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref}
\usepackage[all]{hypcap}

@@

\providecommand{\tabularnewline}{\\}
% Alter some LaTeX defaults for better treatment of figures:
    % See p.105 of "TeX Unbound" for suggested values.
    % See pp. 199-200 of Lamport's "LaTeX" book for details.
    %   General parameters, for ALL pages:
    \renewcommand{\topfraction}{0.9}	% max fraction of floats at top
    \renewcommand{\bottomfraction}{0.8}	% max fraction of floats at bottom
    %   Parameters for TEXT pages (not float pages):
    \setcounter{topnumber}{2}
    \setcounter{bottomnumber}{2}
    \setcounter{totalnumber}{4}     % 2 may work better
    \setcounter{dbltopnumber}{2}    % for 2-column pages
    \renewcommand{\dbltopfraction}{0.9}	% fit big float above 2-col. text
    \renewcommand{\textfraction}{0.07}	% allow minimal text w. figs
    %   Parameters for FLOAT pages (not text pages):
    \renewcommand{\floatpagefraction}{0.7}	% require fuller float pages
	% N.B.: floatpagefraction MUST be less than topfraction !!
    \renewcommand{\dblfloatpagefraction}{0.7}	% require fuller float pages

	% remember to use [htp] or [htpb] for placement

@@

Here we look at the merger history of 13 exemplary galaxies across the galaxy mass spectrum extracted from a cosmological simulation of hierarchical structure formation. We investigate how, after merging with incoming galaxies, SMBHs sink into the cores of the hosts and interact with the resident black hole. We show that gravitational interactions of multiple SMBHs are most probable in high-mass galaxies with total mass $10^{12} \msun < M < 10^{13} \msun$. Galaxies with lower masses have too few mergers with SMBH hosting galaxies. Galaxies with higher masses are more extended, making dynamical friction processes less efficient and hence failing to drive SMBHs into the host galaxy core.

@@

\subsection{Overview of simulation}
Our simulations focus on elliptical galaxies with central SMBHs. The galaxies were given a dark matter background potential using the Stone-Ostriker profile (\citet{2015ApJ...806L..28S}), which is a three-parameter potential-density pair, whose quantities such as density, potential, and binding energy can be written in closed form, having a profile similar to that of a truncated isothermal sphere.  The galaxies were evolved from $4 > z > 0$.  Orbiting black holes were periodically introduced into the "host" galaxy following mergers and their dynamical interaction with the background potential and central SMBH were followed, as described further below.

@@

The primary data we extracted from the above work was organized as follows:
\begin{itemize}
    \item Galactic properties - Each galaxy was distinguished by an id number, with properties such as stellar mass, dark matter mass, central black hole id number, and orbiting black hole id numbers, recorded at 37 redshift slices between $4 > z > 0$.
    \item Black Hole properties - Each black hole was also distinguished by an id number, with properties such as seed mass, accreted mass, its host galaxy id number, and time after $z=4$ at which the black hole entered its host galaxy, if it was not the central black hole.
\end{itemize}

@@

We also compared the stellar masses from Lackner12 to  observational and abundance matching data from \citet{2018AstL...44....8K}. Figure \ref{fig:stellar1} is a partial, approximate, replication of their Figure 11 (abundance matching line).  We rescaled our stellar masses at $z=0$ in a lognormal manner about the line in order to match the observed data.  Table \ref{table:gal_char} shows the original and rescaled stellar masses.

@@

\begin{table}[htbp]
\caption{Galaxy characteristics at z=0}
\begin{tabular} {|c | c | c| c|}
\hline
$M_{DM}$ & $M_{CBH}$ & $M_{*} (orig.)$ & $M_{*} (rescaled)$ \\
\hline
 \num{9.18e13} 	&	 \num{2.19e9} 	&	\num{1.75e13} 	&	\num{5.14e11}  \\
 \num{7.75e13} 	&	 \num{5.27e9}	&	\num{1.52e13} 	&	\num{4.00e11}  \\
 \num{2.85e13} 	&	 \num{1.46e9}	&	\num{5.58e12} 	&	\num{3.30e11}  \\
 \num{2.48e13} 	&	 \num{2.40e9}	&	\num{4.80e12} 	&	\num{2.25e11}  \\
 \num{2.29e13} 	&	 \num{1.22e9}	&	\num{4.46e12} 	&	\num{2.68e11}  \\
 \num{1.18e13} 	&	 \num{3.62e9}	&	\num{2.32e12} 	&	\num{2.83e11}  \\
 \num{1.03e13} 	&	 \num{9.85e9}	&	\num{1.96e12} 	&	\num{1.62e11}  \\
 \num{1.00e13} 	&	 \num{9.32e9}	&	\num{1.95e12} 	&	\num{2.49e11}  \\
 \num{8.92e12} 	&	 \num{5.60e9}	&	\num{1.74e12} 	&	\num{3.20e11}  \\
 \num{8.56e12} 	&	 \num{6.86e9}	&	\num{1.62e12} 	&	\num{1.83e11}  \\
 \num{6.94e12} 	&	 \num{2.02e9}	&	\num{1.72e12} 	&	\num{1.51e11}  \\
 \num{6.29e12} 	&	 \num{1.22e9}	&	\num{1.23e12} 	&	\num{1.87e11}  \\
 \num{1.84e12} 	&	 \num{2.21e9}	&	\num{5.52e11} 	&	\num{6.42e10}  \\
\hline
\end{tabular}

\label{table:gal_char}
\end{table}

@@

At each iteration in the code, $r_h$ is updated according to the stellar mass and Equations \ref{re} and \ref{sig}.  The core radius, $r_c$, is recalculated only if an orbiting black hole gets within $r_c$.  The work done by diffusion, as described in Section \ref{psd} is calculated, the total potential energy and dynamical friction are updated, and $r_c$ is solved for from Equation 8 in Stone-Ostriker.

@@

We can identify Equation \ref{eq:df} as the dynamical friction term. The second term introduces a variance of the friction term, and even allows the SBHs to be accelerated when the velocity of a SBH is sufficiently small. The third term introduces a change in velocity perpendicular to the direction of motion of the SBH. It is a randomly oriented vector, and hence causes the SBHs to execute a random walk in phase space. The last two terms will establish that the SBHs are ultimately in energy equipartition with the background stars.

@@

\subsubsection{Gravitational wave recoils}
The code \textsc{AR-Chain} includes post-Newtonian terms up to order 2.5. The SMBHs can therefore merge via gravitational wave emission. We include gravitational wave recoils following the prescription outlined in Kulier15, which is based on the fitting formula by \citet{2012PhRvD..85h4015L}.  We assume that a merger will be inevitable when the separation between two SBHs becomes smaller than 1.0 Schwarzschild radius. At the moment of the merger, we assume that the spin vectors of the two SBHs are randomly aligned.

@@

\subsection{Simulation setup}
Using the merger tree data, orbiting black holes were injected into its host galaxy at redshift z, at a distance from the galactic center of $R_{e}$ (Eqn. \ref{re}) following galaxy mergers.  Their initial velocity was arbitrarily chosen to be circular, $v_c = \sqrt{M(R_e)/R_e}$, with $v_x$, $v_y$, and $v_z$ randomly chosen.  Note: we only inject the BHs into the simulation that have a $t_{fric}$ smaller than 100 times the Hubble time.

\subsection{Black Hole Luminosity}
The x-ray temperature versus sigma relation from Table 2 of  \citet{2018ApJ...857...32B} was fitted according to:
\begin{equation}
    T_x = 10^{(2.50*log \sigma - 6.06)}
\end{equation}
Central gas density, ${\rho}_0$ was also fitted from Table 3 of \citet{2018ApJ...857...32B} according to:
\begin{equation}
    \rho_0 = 10^{(0.6log(\sigma) - 25)}
\end{equation}
The core radius of the gas model, ${r_c}$ is:
\begin{equation}
    r_c = 10^{(15.09log(\sigma) - 35.77)}
\end{equation}
Local gas density ${\rho}$ is obtained from the isothermal beta model:
\begin{equation}
    \rho = \frac{\rho_0}{(1+r^2/r_c^2)^{1.5}}
\end{equation}
Gas sound speed, ${ss}_g$, was obtained from:
\begin{equation}
    {ss}_g = \sqrt{1.67T_x/m_p},
\end{equation}
where $m_p$ is the proton mass.

The black hole mass accretion rate is approximated by:
\begin{equation}
    \dot{M}_{acc} = \frac{4{\pi}G^2M^2\rho}{({ss_g}^2+v^2)^{3/2}},
\end{equation}
where M is black hole mass, G is the gravitational constant, and v is the black hole velocity.

Finally, the black hole luminosity is calculated from:
\begin{equation}
    L = 0.1\dot{M}_{acc}{ss_g}^2
\end{equation}

The Eddington limit is known to be:
\begin{equation}
    L_{edd} = \num{1.26e38} (M/M_{\odot}) erg/s
\end{equation}
